Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-05 paper/section-general-framework.tex
Better explanation of the construction of relational environments

The definition of relational environments is the most subtle part of
our development. The straightforward approach would be to assume some
interpretation of the indexing expressions, where sorts are
interpreted as sets and index operations as functions, and then define
the relational interpretation of a primtive type
$\tyPrim{X}{e_1,...,e_n}$ in terms of the interpretations of
$e_1,...,e_n$. This approach suffices to prove free theorems and type
isomorphisms for instantiations of our general framework. However, in
order to prove non-definability results (like, for example, the one in
\exref{ex:uninhabited-type}), we choose a more subtle interpretation
of index contexts and primitive types that can account more precisely
for the definable programs permitted by the algebraic indexing. In
this section, we set out the general requirements that we require of
relational environments in order for the abstraction theorem to
hold. In \autoref{sec:constr-rel-env}, we give a general construction
of relational environments that will allow us to prove several
non-definability results.

@@

\subsubsection{A General Construction of Relational Environments}
\label{sec:constr-rel-env}

We now give a general construction of valid families of sets of
relational environments that will allow us to derive useful
consequences of the abstraction theorem, including free theorem, type
isomorphisms and non-definability results. We make use of some
concepts of universal algebra, which we recall here. \fixme{Desperately needs some intuition: do so in terms of clumpiness}

\paragraph{Some Universal Algebra}
For each sort $s \in \SortSet$ we assume a carrier set
$\semSort{s}$. We extend this assignment to index contexts and other
sequences of sorts (e.g., type arities) using the cartesian product of
sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of $e$. We further assume
that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ such that all the axioms in
$\IndexAxiomSet$ are satisfied.

% A sub-model $\mathcal{N}$ of a given
% model $\mathcal{M}$ (written $\mathcal{N} \subseteq \mathcal{M}$) is a
% model such that $\semSort{s}_{\mathcal{N}} \subseteq
% \semSort{s}_{\mathcal{M}}$ for all $s \in \SortSet$.

\fixme{Define free models for any sort-indexed collection of sets}
Given an index context $\Delta$, the \emph{free} model on $\Delta$,
$\Free(\Delta)$, has $\semSort{s}_{\Free(\Delta)} = \{e \sepbar \Delta
\vdash e : s\}/\mathord\equiv$: the set of expressions of sort $s$ in
the indexing context $\Delta$, quotiented by the equality
judgment. The free model has
$\semIndexExp{\texttt{f}}_{\Free(\Delta)}(e_1,..,e_n) =
\texttt{f}(e_1,...,e_n)$.

\paragraph{Relational Environments} \fixme{Say some stuff in english to make sure that the construction is understandable}

To assign a relational meaning to
each of the primitive types we assume a some model $\mathcal{M}$ of
the indexing theory along with a function $R_{\tyPrimNm{X}} :
\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
\Rel(\tyPrimSem{\tyPrimNm{X}})$ for each primitive type
$\tyPrimNm{X}$, and a constant $R^\bullet_{\tyPrimNm{X}} \in
\Rel(\tyPrimSem{\tyPrimNm{X}})$. We set
$\relEnv{E}_{\mathcal{M},R}(\Delta) = \{ \rho^{\mathcal{A},h} \sepbar \mathcal{A}
\subseteq \Free(\Delta), h : \mathcal{A} \to \mathcal{M} \}$, where:
\begin{displaymath}
  \rho^{\mathcal{A},h}\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
    \begin{array}{ll}
      R^\bullet_{\tyPrimNm{X}} & \textrm{if any }e_i \not\in \mathcal{A} \\
      R_{\tyPrimNm{X}}(h(e_1),...,h(e_n)) & \textrm{if all }e_i \in \mathcal{A}
    \end{array}
  \right.
\end{displaymath}

\begin{lemma}
  The family $\relEnv{E}_{\mathcal{M},R}$ of sets of relation
  environments is valid (c.f.~\defref{defn:valid-rel-env-family}).
\end{lemma}

% \fixme{Note that we can recover the basic approach by restricting the
%   sub-algebras to just the whole thing. Does this work in general?
%   What about arbitrary collections of sub-algebras?}

\begin{example}[Two-dimensional Geometry] We define the ``standard''
  model of the indexing theory for the two-dimensional geometry
  example as follows. Each of the sorts are interpreted just as their
  semantic counterparts:
  \begin{displaymath}
    \begin{array}{@{}llll}
      \semSort{\SynTransl{2}} = \Transl{2}
      &
      \semSort{\SynGL{2}} = \GL{2}
      &
      \semSort{\SynOrth{2}} = \Orth{2}
      &
      \semSort{\SynGL{1}} = \GL{1}
    \end{array}
  \end{displaymath}
  Each of the index operations (e.g.,~the group structure and
  determinant) are interpreted by the intended semantic
  operations. These clearly satisfy the axioms we stated in
  \exref{ex:two-dim-geo-axioms}. The assignment of relations for the
  primitive types $\tyPrimNm{vec}$ and $\tyPrimNm{real}$ are given by
  the following definitions:
  \begin{displaymath}
    \begin{array}{ll}
      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
    \end{array}
  \end{displaymath}
  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
  are as may be expected given the informal definitions in
  \autoref{sec:motivating-examples}. We have set
  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ to
  relate $\vec{0}$ and $0$ to themselves to account for the
  polymorphic zero constants we have assumed.
\end{example}

@@

>>>>

The definition of relational environments is the most subtle part of
our development. The straightforward approach would be to assume some
interpretation of the indexing expressions, where sorts are
interpreted as sets and index operations as functions, and then define
the relational interpretation of a primtive type
$\tyPrim{X}{e_1,...,e_n}$ in terms of the interpretations of
$e_1,...,e_n$. This approach suffices to prove free theorems and type
isomorphisms for instantiations of our general framework. However, in
order to prove non-definability results (like, for example, the one in
\exref{ex:uninhabited-type}), we choose a more refined interpretation
that accounts more precisely for the definable programs permitted by
the algebraic indexing. In this section, we set out the general
requirements that we require of relational environments in order for
the abstraction theorem to hold. In \autoref{sec:constr-rel-env}, we
give a general construction of relational environments that will allow
us to prove several non-definability results.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A General Construction of Relational Environments}
\label{sec:constr-rel-env}

We now give a general construction of valid families of sets of
relational environments (in the sense of
\defref{defn:valid-rel-env-family}) that will allow us to derive
useful consequences of \thmref{thm:abstraction}. In particular, we
wish to show that certain types are uninhabited. To motivate this,
consider the following type, which makes use of vectors indexed by
translations, as in \autoref{sec:affine-vector-ops}:
\begin{displaymath}
  \forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
\end{displaymath}
In \exref{ex:uninhabited-type} we claimed that this type is
uninhabited. Intuitively, it is clear why: given the operations on
vectors we presented in \autoref{sec:affine-vector-ops}, there is no
way of cancelling the additional $t$ in the argument type to get the
result type. More generally, the operations on vectors only allow us
to combine indexes according to the group operations. Thus, if the
index of the result type of a function does not lie within the free
abelian group generated by the indexes of the input types, then there
is no way of writing a program with the given type. We now develop
this idea to generate a family of sets of relational environments that
is valid in the sense of \defref{defn:valid-rel-env-family}. The key
idea is to assign different relational interpretations to types with
indexes that lie outside some fintely generated of index expressions.

\paragraph{Some Universal Algebra}
We make use of some concepts of universal algebra, which now recall.

A \emph{model} for our indexing expressions and their equality is
defined as follows. For each sort $s \in \SortSet$ we assume a carrier
set $\semSort{s}$. We extend this assignment to index contexts and
other sequences of sorts (e.g., type arities) using the cartesian
product of sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} =
\semSort{s_1} \times ... \times \semSort{s_n}$. For each index
operation $\texttt{f} \in \IndexOpSet$ with $\indexOpArity(\texttt{f})
= ([s_1,...,s_n],s)$, we assume a function $\semIndexExp{\texttt{f}} :
\semSort{s_1,...,s_n} \to \semSort{s}$. For each well-sorted index
expression $\Delta \vdash e : s$, we assign a function
$\semIndexExp{e} : \semSort{\Delta} \to \semSort{s}$ by recursion on
the structure of $e$. We further assume that for each axiom $\Delta
\vdash e \stackrel{ax}\equiv e' : s \in \IndexAxiomSet$, we have
$\semIndexExp{e} = \semIndexExp{e'}$. A \emph{model} is a pair
$\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ such that all the axioms in
$\IndexAxiomSet$ are satisfied.

Let $A = \{ A_s \}_{s \in \SortSet}$ be a family of finite sets
indexed by sorts. We construct an index context $\Delta_A$ which has
variables $a\mathord:s$ for each $a \in A_s$. The \emph{free} model
over $A$, denoted $\Free(A)$, is defined by setting
$\semSort{s}_{\Free(A)} = \{e \sepbar \Delta_A \vdash e :
s\}/\mathord\equiv$, where we quotient by the equality judgment
defined in \autoref{sec:type-equality}. The free model has
$\semIndexExp{\texttt{f}}_{\Free(A)}(e_1,..,e_n) =
\texttt{f}(e_1,...,e_n)$.

\paragraph{Relational Environments} We now fix some model
$\mathcal{M}$ that will provide the underlying interpretation of index
expressions. As we indicated above, we wish to discriminate between
index expressions that are generated from some finite set of ``input''
expressions and those that are not. Accordingly, we assume that for
each primitive type $\tyPrimNm{X}$ we have two ways of assigning a
relational meaning to that type. When the index arguments of the type
are within the generated set, we will use a function $R_{\tyPrimNm{X}}
\in \semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
\Rel(\tyPrimSem{\tyPrimNm{X}})$ that assigns a relational meaning to
every element in the interpretation of the type's argument sorts. When
the index arguments are not within the generated set we default to a
constant relational meaning $R^\bullet_{\tyPrimNm{X}} \in
\Rel(\tyPrimSem{\tyPrimNm{X}})$.

We now use $R_{\tyPrimNm{X}}$ and $R^\bullet_{\tyPrimNm{X}}$ to
construct the relational environments that we are interested in.  Fix
an index context $\Delta$. Let $A = \{ A_s \}_{s \in \SortSet}$ be a
sort-indexed family of finite sets of expressions such that if $e \in
A_s$, then $\Delta \vdash e : s$. Let $h : \Free(A) \to \mathcal{M}$
be a homomorphism. For each $A$ and $h$ we have a relational
environment for the context $\Delta$:
\begin{displaymath}
  \rho^A_h\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
    \begin{array}{@{}l@{\hspace{0.5em}}l}
      R^\bullet_{\tyPrimNm{X}} & \textrm{if any }e_i \not\in \Free(A) \\
      R_{\tyPrimNm{X}}(h(e_1),...,h(e_n)) & \textrm{if all }e_i \in \Free(A)
    \end{array}
  \right.
\end{displaymath}
Collecting all these relational environments together yields a family
of relational environments, parameterised by our choice of model,
$\mathcal{M}$, and relational interpretations of primitive types
$R^\bullet,R$:
\begin{displaymath}
  \relEnv{E}_{\mathcal{M},R,R^\bullet}(\Delta) = \{ \rho^A_h \sepbar \forall s.A_s \subseteq \semSort{s}_{\Free(\Delta)}, h : \Free(A) \to \mathcal{M} \}
\end{displaymath}

\begin{theorem}
  The family $\relEnv{E}_{\mathcal{M},R,R^\bullet}$ of sets of
  relation environments is valid in the sense of
  \defref{defn:valid-rel-env-family}.
\end{theorem}

Note that we still need to prove for each instantiation of our general
framework that the program-level primitive operations preserve all the
relations, in order to fulfil the hypothesis of
\thmref{thm:abstraction}. \fixme{This is the content of lemma X, lemma
  Y, lemma Z, below}.
